Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-27 paper.qmd
rewrite, missing conclusion

The most comprehensive approach to date is containerisation with [Docker]{.pkg}.
[Docker]{.pkg} packages a *data product* together with all its
dependencies—including the operating system, system libraries, programming
languages, and packages—into a self-contained image. Once built, the analysis
can be executed inside a *container*, a running instance of the image.

The strength of [Docker]{.pkg} lies in its ability to bundle not only the
correct versions of [R]{.proglang} and its packages but also system-level
dependencies. Since [Docker]{.pkg} images are minimal Linux systems, they
include the libraries required by packages such as [sf]{.pkg}, ensuring that
future replicators have access to the same environment. Moreover, these images
can be shared via registries like Docker Hub, enabling reproducibility across
systems.

The Rocker project [@boettiger2017] provides pre-built [Docker]{.pkg} images for
the [R]{.proglang} community, offering specific R versions and, in some cases,
preinstalled packages. These serve as convenient bases for reproducible analysis
environments.

Despite its strengths, [Docker]{.pkg} presents several challenges:

1. Poor interactive support: While containers can be modified at runtime,
   changes are lost unless explicitly saved. Running graphical applications like
   RStudio Desktop requires complex X11 configuration that works reliably only
   on Linux. Web-based IDEs such as RStudio Server are alternatives but have
   their own limitations.

2. Steep learning curve: Writing reliable `Dockerfile`s requires familiarity
   with Linux system administration. To ensure reproducibility, `Dockerfile`s
   should reference image *digests* rather than *tags*, but this is rarely done.
   Packages like [dockerfiler]{.pkg} [@dockerfiler] help automate this but do
   not remove the need for basic Linux knowledge.

3. Post-hoc reproducibility: Analyses are often developed interactively and
   containerised only afterward, achieving post-hoc reproducibility rather than
   consistent environments during development.

4. Single-language focus: While [Docker]{.pkg} can support polyglot setups,
   orchestrating them requires manual coordination. There is no standard way to
   specify that one step runs in [Python]{.proglang} and another in [R]{.proglang}.

5. Incomplete reproducibility: Reproducibility is best viewed as a spectrum
   rather than a binary property [@dellaieraMasterThesis2024]. [Docker]{.pkg}
   achieves strong run-time reproducibility—consistency across systems
   (space)—but weak build-time reproducibility, or consistency over time.
   Mutable base images and non-deterministic commands such as `apt-get update`
   introduce temporal drift: rebuilding the same `Dockerfile` at different times
   can yield different results. Even with pinned base images, container
   checksums can vary across builds and machines. Thus, while [Docker]{.pkg}
   provides strong run-time reproducibility, achieving deterministic artefacts
   requires pinning image digests, freezing dependencies, and using
   deterministic build systems such as [Nix]{.pkg} for full control over both
   time and space [@Malka_2024].

Despite these shortcomings, [Docker]{.pkg} remains valuable for specific use
cases: production deployment, CI/CD integration, and cross-platform
distribution. The two approaches are not mutually exclusive. [Nix]{.pkg} can be
installed and used inside [Docker]{.pkg}, combining [Nix]{.pkg}'s deterministic
builds with [Docker]{.pkg}'s deployment infrastructure. In this workflow,
[Nix]{.pkg} handles environment reproducibility during the build of the image,
while [Docker]{.pkg} serves as the distribution mechanism. This addresses
[Docker]{.pkg}'s build-time reproducibility issues while maintaining its deployment
advantages. However, for research workflows where interactive development is
primary and [Docker]{.pkg}'s complexity is a barrier, [Nix]{.pkg} alone may suffice.

Thankfully, it is entirely possible to combine both [Docker]{.pkg} and [Nix]{.pkg}:
[Nix]{.pkg} to setup reproducible environments and [Docker]{.pkg} for deployment
across the many different already existing infrastructures.

Even with a perfectly reproducible environment, researchers face another
challenge: reliably executing the analysis workflow. Which scripts run first?
What are the dependencies between steps? Manual execution is error-prone and
poorly documented. Build automation tools like [Make]{.pkg} address this by
defining analyses as ordered, reproducible steps.

@@

@mcdermott2021 provides an excellent example of best practice. The accompanying
repository^[See
[https://github.com/grantmcdermott/skeptic-priors](https://github.com/grantmcdermott/skeptic-priors)]
demonstrates:

- Package versions recorded in an `renv.lock` file;
- A `Makefile` automating the full analysis;
- A `Dockerfile` providing the computational environment.

Achieving this required mastering multiple complex tools: [renv]{.pkg} for
package management, [Docker]{.pkg} for containerisation, and [Make]{.pkg} for
orchestration. The complexity increases for polyglot projects, which must manage
multiple ecosystems simultaneously.

@@

## The Nix package manager {#sec-nix}

[Nix]{.pkg} is a powerful, cross-platform package manager designed for
installing and building software in a fully reproducible and reliable manner.
Unlike traditional package managers, [Nix]{.pkg} emphasises immutable
infrastructure and a functional approach, which significantly enhances the
consistency of computational environments. As of writing, the primary [Nix]{.pkg}
package collection, `nixpkgs`, provides access to over 120,000 packages,
including nearly all of CRAN and Bioconductor. According to Repology, a service
that tracks package repositories across distributions, `nixpkgs` is both the
largest and among the most up-to-date package repositories available (see Figure
@fig-repology).

@@

This extensive coverage allows users to install not only [R]{.proglang} itself
but also all its required packages and system-level dependencies for any given
project.

While [Nix]{.pkg} is the default package manager for the NixOS Linux
distribution, it can be installed as a standalone tool on other Linux
distributions, macOS, and effectively on Windows via the Windows Subsystem for
Linux (WSL2), treating these as equivalent platforms in practice.

A key advantage of using [Nix]{.pkg} to install [R]{.proglang} packages, as
opposed to the standard `install.packages()` function, lies in its comprehensive
dependency management. [Nix]{.pkg} ensures that all dependencies of each package
are installed, irrespective of whether they are other [R]{.proglang} packages or
underlying system libraries.

Consider, for example, the [sf]{.pkg} package, which is fundamental for spatial
data analysis in [R]{.proglang}. [sf]{.pkg} relies on several complex external
system libraries such as [GDAL]{.pkg}, [GEOS]{.pkg}, and
[PROJ]{.pkg}. Manually installing and configuring these libraries to be
compatible with [sf]{.pkg} can be challenging and platform-specific. With
[Nix]{.pkg}, however, the user merely needs to declare [sf]{.pkg} as a project
requirement. [Nix]{.pkg} then automatically installs and configures all
necessary system libraries as transitive dependencies. This seamless process is
possible because the maintainers of the [R]{.proglang} programming language for
[Nix]{.pkg} have declaratively specified these dependencies for [sf]{.pkg} in
`nixpkgs`, allowing the entire software stack to be provisioned effortlessly
from an end-user perspective.

This concept is central to [Nix]{.pkg}'s design, where packages are referred to
as *component closures*. As @dolstra2004nix explains:

> The idea is to always deploy component closures: if we deploy a component,
> then we must also deploy its dependencies, their dependencies, and so on. That
> is, we must always deploy a set of components that is closed under the
> ''depends on'' relation. Since closures are selfcontained, they are the units
> of complete software deployment. After all, if a set of components is not
> closed, it is not safe to deploy, since using them might cause other
> components to be referenced that are missing on the target system.

The core of [Nix]{.pkg}'s reproducibility lies in its functional package
management paradigm and its unique approach to builds. When installing software
with [Nix]{.pkg}, it evaluates an *expression* written in the [Nix]{.proglang}
language. These expressions define *derivations*, which are declarative
blueprints describing how to build a package. A derivation specifies all inputs:
source code, build commands, and crucially, all direct and transitive
dependencies. This declarative nature ensures that every build process is fully
specified and independent of the environment where it's executed. Furthermore,
all [Nix]{.pkg} expressions for official packages are hosted in the `nixpkgs`
GitHub repository. By referencing a specific commit of `nixpkgs` (a process
known as *pinning a revision*) users guarantee that all packages installed are
derived from the exact same set of build instructions. This means that if an
environment is built today using a pinned revision, it will produce the
identical set of software versions and configurations if rebuilt tomorrow, next
year, or on a different machine, as long as the `nixpkgs` repository remains
accessible.

Pinning is essential, but it is not the only mechanism through which [Nix]{.pkg}
achieves reproducibility. [Nix]{.pkg} adopts principles from functional
programming, conceptualising software builds as *pure functions*. In
mathematics, a pure function is defined as one that always produces the same
output for a given input—for instance, if \( f(x) = x^2 \), then \( f(2) = 4 \)
invariably holds. Analogously, [Nix]{.pkg} ensures that, given identical inputs
(such as source code, dependencies, and build instructions), the resulting
output (the built package) remains identical, irrespective of the environment or
time of execution. In this sense, [Nix]{.pkg} functions as a *functional package
manager*, applying the deterministic properties of pure functions to software
construction.

This is achieved by disallowing side effects and ensuring that builds operate
within hermetic (isolated) environments, with no hidden dependencies on global
system state. The principles of functional package management (hermetic builds,
immutable artifacts, and declarative dependencies) are designed to produce
deterministic outputs. The effectiveness of this model was recently validated in
the most extensive study of build reproducibility to date. The effectiveness of
this model was recently validated at unprecedented scale by @malka2025, who
rebuilt 709,816 packages from historical nixpkgs snapshots (2017–2023). They
found rebuildability rates exceeding 99% and bitwise reproducibility rates
between 69% and 91%, with an upward trend over time. Importantly, they
identified that about 15% of unreproducible builds failed due to embedded build
timestamps, which a solvable issue rather than a fundamental limitation.

While this functional approach greatly enhances reproducibility, it can
sometimes introduce complexity for package maintainers, especially for software
that needs to download external assets during installation (e.g., certain
Bioconductor packages). For end-users, however, this complexity is largely
abstracted away thanks to both [rix]{.pkg} and [rixpress]{.pkg}.

Additionally, [Nix]{.pkg} supports multiple versions or *variants* of a package
on the same system. [Nix]{.pkg} installs all packages into a centralised,
content-addressed directory called the [Nix]{.pkg} store (typically
`/nix/store`). Each package gets a unique subdirectory whose name includes a
hash of all its inputs: source code, dependencies, build instructions. This
means `/nix/store/abc123-dplyr-1.0.0` and `/nix/store/xyz789-dplyr-1.1.0` can
coexist without conflicts.

This isolation ensures that projects always use their intended software
versions, allowing multiple versions of [R]{.proglang} to coexist and preventing
"dependency hell" and global environment pollution.

@@

By leveraging these principles, [Nix]{.pkg} can effectively replace and
integrate the functionalities of tools like [renv]{.pkg} and [Docker]{.pkg} for
[R]{.proglang} projects, or `requirements.txt` files and virtual environments
for [Python]{.proglang} projects. Furthermore, [Nix]{.pkg} excels at building
polyglot environments, which can seamlessly combine [R]{.proglang},
[Python]{.proglang}, [Julia]{.proglang}, a [LaTeX]{.proglang} distribution, or
any of the numerous other tools available in `nixpkgs`. This enables the
creation of a truly complete, project-specific, and deeply reproducible
environment that can be used interactively for development or non-interactively
for automated analysis. As long as the `nixpkgs` repository (or a suitable fork,
as discussed in Section @sec-fork) remains accessible, the environment can be
rebuilt reliably in the future.

[Nix]{.pkg} provides strong cross-platform support across major operating
systems. It runs natively on Linux (x86_64) with full functionality and optimal
integration. On macOS (Intel and Apple Silicon), however, reproducibility is
less reliable due to dependencies on impure system frameworks and Xcode
toolchains outside the Nix store. As a result, macOS users may need to update
project pins after system or Xcode upgrades to restore builds. On Windows,
[Nix]{.pkg} works well through the Windows Subsystem for Linux (WSL2), though
graphical applications may need extra setup. [Nix]{.pkg} environments on WSL can
be used interactively via VS Code or Positron for a smooth development
experience. It is also important to note that [Nix]{.pkg} can be installed
within a [Docker]{.pkg} image to deploy and reproduce environments reliably in
containerized workflows.

While [Nix]{.pkg} is a powerful system, it comes with notable challenges. The
steepest is its learning curve: [Nix]{.pkg} uses its own declarative language
(also called [Nix]{.proglang}), which can be difficult for those new to
functional or declarative paradigms. Complexity becomes most apparent when
writing custom derivations or troubleshooting intricate build issues. Build
times can also be long, especially for large environments, when pre-built
binaries aren’t available. Additionally, because [Nix]{.pkg} stores all package
outputs immutably in the Nix store, disk usage tends to be higher than with
traditional package managers. Finally, while `nixpkgs` is extensive, some new or
niche packages may be missing or fail to build across platforms.

@@

This expression can be saved in a file called `default.nix`. The environment can
then be built on a system with [Nix]{.pkg} installed using the `nix-build`
command.^[For installing [Nix]{.pkg}, we recommend the Determinate Systems
installer: \url{https://determinate.systems/posts/determinate-nix-installer}].
Once the build completes, the user can enter the interactive shell with
`nix-shell`, which executes following steps:

- Nix reads the expression and identifies what needs to be installed;
- It checks if these packages already exist in `/nix/store`;
- If not, it downloads or builds them (with all dependencies);
- It creates a new shell environment and updates the `$PATH` to include the
  right binaries from `/nix/store`.
- The user is now in an interactive [Nix]{.pkg} shell session.

@@

[rix]{.pkg} provides the `rix()` function, which simplifies generating
[Nix]{.pkg} expressions. It is available on CRAN and can be installed like any
other R package. Additionally, it can bootstrap an [R]{.proglang} development
environment on a system where [R]{.proglang} is not yet installed but
[Nix]{.pkg} is available. This can be done by running (inside of a terminal):

@@

This call to `rix()` generates a `default.nix` file for a development shell that
encapsulates a complete and reproducible research environment. It provides
[R]{.proglang} and its packages (including [AER]{.pkg} at version 1.2-8),
several TeXLive packages for \LaTeX document authoring, development versions of
[rix]{.pkg} and [fusen]{.pkg} pulled directly from GitHub at a specific commit,
and the Positron editor. The reproducibility of this environment is guaranteed
by pinning all components to a single point in time: the [R]{.proglang} packages
are resolved from the CRAN snapshot of October 20, 2025, while all other system
tools are fixed to the version of `nixpkgs` from that same date. Typing
`nix-shell` in a terminal within the folder that contains this `default.nix`
will drop the user into a new shell. It should be noted that it uses the
`nixpkgs` from that same date. It is also possible to configure IDEs to
dynamically load this new shell to provide the proper development tooling.

@@

[Python]{.proglang} packages, however, require special handling. Unlike CRAN or
the Julia package registry, PyPI (the [Python]{.proglang} Package Index)
contains over 500,000 packages with highly variable quality—many lack proper
build specifications, have broken dependencies, or are abandoned. Because of
this heterogeneity, [Nix]{.pkg} maintainers cannot automatically package all of
PyPI. Instead, they manually curate and package individual [Python]{.proglang}
packages that meet quality standards. This means some [Python]{.proglang}
packages, or specific versions, may not be available through [Nix]{.pkg}.

For projects requiring [Python]{.proglang} packages not yet available in
[Nix]{.pkg}, [rix]{.pkg} supports integration with [uv]{.pkg}, a modern
[Python]{.proglang} package manager. This allows users to combine Nix's
system-level reproducibility ([Python]{.proglang} interpreter, system libraries)
with [uv]{.pkg}'s comprehensive package availability. While this hybrid approach
sacrifices some of Nix's deterministic build guarantees for [Python]{.proglang}
packages, it provides a pragmatic solution when the needed packages are
unavailable in `nixpkgs`.

@@

While this approach doesn't achieve Nix's level of build-time determinism for
[Python]{.proglang} packages ([uv]{.pkg} still downloads wheels or builds
packages at install time rather than using pre-built, content-addressed
artefacts) it provides a pragmatic balance: comprehensive [Python]{.proglang}
package availability with reasonable reproducibility guarantees. For polyglot
projects, this means [R]{.proglang} and system-level dependencies remain fully
reproducible through [Nix]{.pkg}, while [Python]{.proglang} packages gain the
reproducibility that [uv]{.pkg} can provide.

[rix]{.pkg} can generate [Nix]{.pkg} expressions even if [Nix]{.pkg} is not
installed on the system. This is useful for continuous integration and
continuous deployment (CI/CD) workflows on platforms such as GitHub Actions. For
instance, the repository containing the source code for this
article^[https://github.com/b-rodrigues/rix_paper] uses GitHub Actions to
compile the paper. Each time a push is made to the master branch, a runner
installs [Nix]{.pkg}, generates the environment from the hosted `default.nix`
file, and compiles the paper using [Quarto]{.pkg} within the reproducible
environment. This ensure that *exactly* the same environment is used on the
author's computer and on the CI/CD without any additional, platform-specific,
configuration.

@@

Using a fork offers several advantages. First, it provides flexibility that the
official `nixpkgs` repository cannot always accommodate. [Nix]{.pkg} is
primarily the package manager for the NixOS Linux distribution, and governance
and technical choices made upstream can limit what [rix]{.pkg} aims to provide.

@@

The `rstats-on-nix` fork allows me to circumvent these limitations. For example,
it provides daily snapshots of CRAN. Each day, the [R]{.proglang} package set is
updated and committed to a dated branch using GitHub Actions. Users can select a
specific date with:

@@

We strive to provide an available date per week: each Monday, a GitHub Action
tests popular packages on Linux and macOS, and only if all tests succeed is the
date added to the list of available dates in [rix]{.pkg}. This ensures users can
reliably install packages, and allows me to backport fixes if needed. For
example, when RStudio was temporarily broken due to a dependency issue
(`boost`), a pull request was submitted to the official `nixpkgs` repository. I
backported the fix to the `rstats-on-nix` fork, making RStudio available to
users of [rix]{.pkg} earlier than upstream, as merging PRs in the official
repository can take some time.

>>>>

The most comprehensive approach to managing a complete computational environment
to date is containerization, for which [Docker]{.pkg} has become the de facto
standard. Docker's principal strength lies in its capacity to package an entire
data product—including the operating system, all system-level libraries,
programming language runtimes, and packages—into a single, self-contained
artifact known as an image. Once this image is built, the analysis can be
executed within a container, which is a running instance of that image. This
methodology effectively solves the problem of missing system-level dependencies;
because Docker images are typically minimal Linux systems, they can bundle the
specific versions of libraries like [GDAL]{.pkg} or [GEOS]{.pkg} required by
packages such as [sf]{.pkg}. These images can then be shared via public
registries, enabling a high degree of reproducibility across different host
systems. The Rocker project, for instance, provides invaluable pre-built Docker
images tailored for the [R]{.proglang} community, offering a convenient
foundation for reproducible research environments [@boettiger2017].

However, while Docker excels at providing this form of spatial reproducibility,
its application in interactive research workflows reveals several significant
limitations. First, the container model is often poorly aligned with the
iterative, interactive nature of data analysis. Running graphical applications
like RStudio Desktop requires complex and platform-specific configuration, while
ephemeral container filesystems mean that changes made during a session are lost
by default. Consequently, a common workflow involves developing an analysis
interactively on a host system and only containerizing it post-hoc, which
achieves reproducibility as an afterthought rather than as a continuous
guarantee during the research process. Packages like @dockerfiler are specifically
made to enable this type of post-hoc reproducibility.

Second, the effective use of [Docker]{.pkg} introduces a considerable learning
curve. Authoring a reliable and truly reproducible `Dockerfile` requires a
degree of familiarity with Linux system administration. Best practices, such as
referencing base images by their immutable content digests rather than mutable
tags, are necessary for long-term stability but are not widely adopted.
Furthermore, while [Docker]{.pkg} can house polyglot environments, it offers no native
mechanism for orchestrating a pipeline where distinct steps might require
different, and potentially conflicting, sets of dependencies. The entire
analysis is typically confined to a single, monolithic environment.

The most critical limitation, however, lies in its guarantees of temporal
reproducibility. As noted by @dellaieraMasterThesis2024, reproducibility is best
viewed as a spectrum. Docker provides strong run-time reproducibility
(consistency across space), but offers weaker build-time reproducibility
(consistency over time). The imperative commands within a `Dockerfile`, such as
`apt-get update`, are non-deterministic by nature and can introduce temporal
drift, where rebuilding the same file at different times yields different binary
artifacts. As empirically demonstrated by @malka2025, achieving deterministic
builds requires a system designed for this purpose, a role for which functional
package managers like [Nix]{.pkg} are architecturally better suited.^[This
contrast between the procedural, imperative commands of a `Dockerfile` and the
declarative, functional model of [Nix]{.pkg} highlights a fundamental dichotomy
in approaches to reproducibility, a concept we will examine in detail as we
introduce the [Nix]{.pkg} package manager in Section @sec-nix.]

This is not to say that [Docker]{.pkg} has no place in a reproducible research
workflow. Its utility for production deployment, continuous integration, and
cross-platform distribution is well-established. Indeed, the two approaches are
not mutually exclusive. As shall be explained in @sec-nix, it is possible
to use [Nix]{.pkg} within [Docker]{.pkg} to benefit from the advantages of
both tools.

Making sure the development environment is reproducible is but one piece of the
puzzle, though. Even with a perfectly reproducible environment, researchers face
another challenge: reliably executing the analysis workflow. Which scripts run
first? What are the dependencies between steps? Manual execution is error-prone
and poorly documented. Build automation tools like [Make]{.pkg} address this by
defining analyses as ordered, reproducible steps.

@@

The current state-of-the-art in reproducible research exemplifies the complexity
of navigating this fragmented landscape. The work of @mcdermott2021 provides an
excellent case study in best practices, with an accompanying repository that is
fully reproducible.^[See
[https://github.com/grantmcdermott/skeptic-priors](https://github.com/grantmcdermott/skeptic-priors)]
Achieving this gold standard, however, required the
masterful integration of multiple complex and disparate tools: [renv]{.pkg} for
managing [R]{.proglang} packages, [Docker]{.pkg} for containerizing the system
environment, and [Make]{.pkg} for orchestrating the analysis. This demonstrates
that while a high degree of reproducibility is attainable, it requires a heroic
effort, forcing researchers to become experts in a brittle and complex
toolchain.

Furthermore, the complexity increases for polyglot projects, which must manage
multiple ecosystems simultaneously.

@@

## The Nix Package Manager {#sec-nix}

[Nix]{.pkg} is a cross-platform package manager designed for reproducible
software installation and deployment. Unlike traditional package managers,
[Nix]{.pkg} emphasizes immutable infrastructure and a functional programming
paradigm to ensure computational environment consistency. The primary
[Nix]{.pkg} package collection, `nixpkgs`, provides access to over 120,000
packages, including nearly all of CRAN and Bioconductor. According to Repology,
a service tracking package repositories across distributions, `nixpkgs` is both
the largest and among the most up-to-date repositories available (Figure
@fig-repology).

@@

This extensive coverage allows users to install not only [R]{.proglang}
but also all required packages and system-level dependencies for any given
project.

A key advantage of [Nix]{.pkg} over traditional installation methods lies in its
comprehensive dependency management. Consider the [sf]{.pkg} package for spatial
data analysis in [R]{.proglang}. It requires several complex system libraries:
[GDAL]{.pkg}, [GEOS]{.pkg}, and [PROJ]{.pkg}. Manually installing and
configuring these libraries is challenging and platform-specific. With
[Nix]{.pkg}, users simply declare [sf]{.pkg} as a project requirement, and
[Nix]{.pkg} automatically installs and configures all necessary system libraries
as transitive dependencies.

This seamless process stems from [Nix]{.pkg}'s concept of *component closures*.
As @dolstra2004nix explains:

> The idea is to always deploy component closures: if we deploy a component,
> then we must also deploy its dependencies, their dependencies, and so on.
> [...] Since closures are self-contained, they are the units of complete
> software deployment.

In other words, every package deployment includes everything it needs to run,
with no hidden dependencies on the host system.

While comprehensive dependency tracking could theoretically be implemented in
any package manager, what distinguishes [Nix]{.pkg} (alongside its cousin
[Guix]{.pkg}) is its *functional package management* paradigm. Traditional
package managers require imperative, step-by-step installation instructions: *do
this, then this, then that*. [Nix]{.pkg} inverts this model; users declare the
desired end state (*I require an environment with these packages*), and
[Nix]{.pkg} determines how to achieve it.

This is possible because [Nix]{.pkg} packages are defined as *derivations*:
declarative blueprints specifying all inputs (source code, build commands,
dependencies) needed to build a package. When installing software, [Nix]{.pkg}
evaluates expressions written in the [Nix]{.proglang} language to produce these
derivations. The declarative nature ensures every build process is fully
specified and independent of the execution environment.

For example, the [rJava]{.pkg} package definition includes the correct
`JAVA_HOME` configuration pointing to the [jdk]{.pkg} dependency. Outside
[Nix]{.pkg}, users must manually set this variable if automatic detection fails.
With [Nix]{.pkg}, the package maintainers handle this complexity once (by
writing the correct [Nix]{.pkg} expressions that perform the installation
purely), and all users benefit.

[Nix]{.pkg} conceptualizes software builds as *pure functions* from functional
programming. In mathematics, a pure function always produces the same output for
a given input: if $f(x) = x^2$, then $f(2) = 4$ invariably holds. Analogously,
[Nix]{.pkg} ensures that identical inputs (source code, dependencies, build
instructions) produce identical outputs (built packages), regardless of when or
where the build occurs.

This determinism is achieved through hermetic (isolated) builds with no hidden
dependencies on global system state nor undeclared side effects. The
effectiveness of this model was validated at unprecedented scale by @malka2025,
who rebuilt 709,816 packages from historical `nixpkgs` snapshots (2017–2023),
finding rebuildability rates exceeding 99% and bitwise reproducibility rates
between 69% and 91%, with an upward trend. Importantly, about 15% of
unreproducible builds failed due to embedded timestamps—a solvable
implementation issue rather than a fundamental model limitation.

All [Nix]{.pkg} package definitions are hosted in the `nixpkgs` GitHub
repository. By *pinning* to a specific commit (or equivalently, a specific
date), users guarantee that all installed packages derive from identical build
instructions. An environment built today using a pinned revision will produce
the same software versions and configurations if rebuilt years later, provided
the `nixpkgs` repository remains accessible (it is archived and distributed
across multiple mirrors).

[Nix]{.pkg} installs all packages into a content-addressed directory called the
Nix store (typically `/nix/store`). Each package occupies a unique subdirectory
whose name includes a cryptographic hash of all its inputs: source code,
dependencies, and build instructions. This means `/nix/store/abc123-dplyr-1.0.0`
and `/nix/store/xyz789-dplyr-1.1.0` can coexist without conflicts, allowing
multiple [R]{.proglang} versions on the same system and eliminating "dependency
hell."

While this functional approach enhances reproducibility, it can introduce
complexity for package maintainers, particularly for software downloading
external assets during installation (e.g., certain Bioconductor packages). For
end-users, however, this complexity is largely abstracted away by [rix]{.pkg}
and [rixpress]{.pkg}, which we introduce in the following sections.

@@

By leveraging these principles, [Nix]{.pkg} provides a unified framework that
can subsume the functionalities of disparate tools. For [R]{.proglang} projects,
it can replace the combination of [renv]{.pkg} for package management and
[Docker]{.pkg} for system-level isolation. Similarly, for [Python]{.proglang}
projects, it offers a more comprehensive alternative to `requirements.txt` files
and virtual environments. Furthermore, [Nix]{.pkg} excels at composing polyglot
environments, seamlessly integrating [R]{.proglang}, [Python]{.proglang},
[Julia]{.proglang}, a [LaTeX]{.proglang} distribution, and any of the numerous
other tools available in `nixpkgs`. This enables the creation of a truly
complete, project-specific, and deeply reproducible environment suitable for
both interactive development and non-interactive, automated analysis. The
long-term viability of this reproducibility is contingent only on the continued
accessibility of the `nixpkgs` repository or a suitable fork, as will be
discussed in Section @sec-fork.

[Nix]{.pkg} offers robust cross-platform support, running natively on Linux
(x86_64) with optimal functionality. On macOS (Intel and Apple Silicon),
however, its reproducibility guarantees are somewhat compromised by dependencies
on proprietary system frameworks and Xcode toolchains that reside outside
of the [Nix]{.pkg} store; consequently, macOS users may need to update project
pins following system upgrades. For Windows users, [Nix]{.pkg} operates
effectively through the Windows Subsystem for Linux (WSL2), where environments
can be integrated into development workflows using editors like VS Code or
Positron.

While [Nix]{.pkg} is a powerful system, its adoption is not without notable
challenges. The most significant is its steep learning curve; it uses its own
programming language, also called [Nix]{.proglang}, which is purely functional, a
paradigm that may be unfamiliar to many researchers. This complexity is most
apparent when authoring custom build instructions or troubleshooting intricate
dependency issues. Additionally, build times can be substantial for large
environments when pre-built binary artifacts are not available from a cache. The
immutable, content-addressed nature of the [Nix]{.pkg} store also leads to
higher disk usage compared to traditional package managers. Finally, while the
`nixpkgs` repository is extensive, certain niche packages may be unavailable or
fail to build on all platforms.

Beyond these technical aspects, a practical hurdle to adoption is the ubiquity
of [Docker]{.pkg}. Given the extensive infrastructure and community knowledge
built around containerization, [Nix]{.pkg} can be perceived as a competing
rather than a complementary technology. This perception, however, is a false
dichotomy. [Docker]{.pkg} and [Nix]{.pkg} allow researchers to solve similar
issues but these tools are not alternatives of one another. Thus, a powerful
hybrid model involves using [Nix]{.pkg} to deterministically build the
environment *during the construction of a [Docker]{.pkg} image*, which is then
deployed using standard container infrastructure. This approach leverages the
strengths of both tools: [Nix]{.pkg} provides robust build-time reproducibility,
while [Docker]{.pkg} serves as a universal distribution and deployment
mechanism. For many research workflows, however, where interactive development
is paramount and the overhead of containerization is a barrier, the unified
model provided by [Nix]{.pkg} alone can provide a more direct and streamlined
path to reproducibility.

@@

This expression can be saved in a file called `default.nix`. The environment can
then be built on a system with [Nix]{.pkg} installed using the `nix-build`
command.^[For installing [Nix]{.pkg}, we recommend the Determinate Systems
installer: \url{https://determinate.systems/posts/determinate-nix-installer}].
Once the build completes, the user can enter the interactive shell with
`nix-shell`, which executes the following steps:

- [Nix]{.pkg} reads the expression and identifies what needs to be installed;
- It checks if these packages already exist in `/nix/store`;
- If not, it downloads or builds them (with all dependencies);
- It creates a new shell environment and updates the `$PATH` to include the
  right binaries from `/nix/store`.
- The user is now in an interactive [Nix]{.pkg} shell session.

@@

[rix]{.pkg}, an [R]{.proglang} package, provides the `rix()` function, which
simplifies generating [Nix]{.pkg} expressions. It is available on CRAN and can
be installed like any other [R]{.proglang} package. Additionally, it can
bootstrap an [R]{.proglang} development environment on a system where
[R]{.proglang} is not yet installed but [Nix]{.pkg} is available. This can be
done by running (inside of a terminal):

@@

This call to `rix()` generates a `default.nix` file for a development shell that
encapsulates a complete and reproducible research environment. It provides
[R]{.proglang} and its packages (including [AER]{.pkg} at version 1.2-8),
several TeXLive packages for \LaTeX document authoring, development versions of
[rix]{.pkg} and [fusen]{.pkg} pulled directly from GitHub at a specific commit,
and the Positron editor. The reproducibility of this environment is guaranteed
by pinning all components to a single point in time: the [R]{.proglang} packages
are resolved from the CRAN snapshot of October 20th 2025, while all other system
tools are fixed to the version of `nixpkgs` from that same date. Typing
`nix-shell` in a terminal within the folder that contains this `default.nix`
will drop the user into a new shell. It is also possible to configure IDEs to
dynamically load this new shell to provide the proper development tooling for
better interactive use.

@@

[Python]{.proglang} packages, however, require special handling. Unlike CRAN or
the Julia package registry, PyPI (the [Python]{.proglang} Package Index)
contains over 500,000 packages with highly variable quality—many lack proper
build specifications, have broken dependencies, or are abandoned. Because of
this heterogeneity, [Nix]{.pkg} maintainers cannot automatically package all of
PyPI as they do for CRAN or the [Julia]{.proglang} registry. Instead, they
manually curate and package individual [Python]{.proglang} packages that meet
quality standards. This means some [Python]{.proglang} packages, or specific
versions, may not be available through [Nix]{.pkg}.

For projects requiring [Python]{.proglang} packages not available in
[Nix]{.pkg}, [rix]{.pkg} supports integration with [uv]{.pkg}, a modern
[Python]{.proglang} package manager. This allows users to combine Nix's
system-level reproducibility ([Python]{.proglang} interpreter, system libraries)
with [uv]{.pkg}'s comprehensive package availability. While this hybrid approach
sacrifices some of Nix's deterministic build guarantees for [Python]{.proglang}
packages, it provides a pragmatic solution when the needed packages are
unavailable in `nixpkgs`.

@@

While this approach doesn't achieve Nix's level of build-time determinism for
[Python]{.proglang} packages ([uv]{.pkg} still downloads wheels or builds
packages at install time rather than using pre-built, content-addressed
artefacts) it provides a pragmatic balance: comprehensive [Python]{.proglang}
package availability with reasonable reproducibility guarantees. For polyglot
projects, this means [R]{.proglang} and system-level dependencies remain fully
reproducible through [Nix]{.pkg}, while [Python]{.proglang} packages gain the
reproducibility that [uv]{.pkg} can provide. We recommend of course to use
[Python]{.pkg} through [Nix]{.pkg} whenever possible and only resort to the
hybrid approach with [uv]{.pkg} as a last resort.^[It is also possible to package
the missing [Python]{.pkg} for [Nix]{.pkg} and submit a pull request to `nixpkgs`,
but we recognise that this option may not be beginner-friendly.]

Anywhere [Nix]{.pkg} can be installed, users can benefit from its features. For
instance, the repository containing the source code for this
article^[https://github.com/b-rodrigues/rix_paper] uses GitHub Actions to
compile the paper. Each time a push is made to the master branch, a runner
installs [Nix]{.pkg}, generates the environment from the hosted `default.nix`
file, and compiles the paper using [Quarto]{.pkg} within the reproducible
environment. This ensures that *exactly* the same environment is used on the
authors' machines and on the CI/CD platform without any additional,
platform-specific, configuration.

@@

Using a fork offers several advantages. First, it provides flexibility that the
official `nixpkgs` repository cannot always accommodate. Because [Nix]{.pkg}
serves as the package manager for the NixOS Linux distribution, upstream
priorities center on system-level stability rather than rapid updates to
language ecosystems like [R]{.proglang} or [Julia]{.proglang}.

@@

The `rstats-on-nix` fork allows us to circumvent these limitations. For example,
it provides daily snapshots of CRAN and the [Julia]{.proglang}. Each day, the
[R]{.proglang} package set is updated and committed to a dated branch using
GitHub Actions. Users can select a specific date with:

@@

We strive to provide an available date per week: each Monday, a GitHub Action
tests popular [R]{.proglang} packages on Linux and macOS, and only if all tests
succeed is the date added to the list of available dates in [rix]{.pkg}. Users
can see all available dates with `rix::available_dates()`. This ensures users
can reliably install packages, and allows us to backport fixes if needed. For
example, when RStudio was temporarily broken due to a dependency issue
(`boost`), a pull request was submitted to the official `nixpkgs` repository. I
backported the fix to the `rstats-on-nix` fork, making RStudio available to
users of [rix]{.pkg} earlier than upstream, as merging PRs in the official
repository can take some time.

We have backported fixes to the `rstats-on-nix` `nixpkgs` fork as far back as
March 2019. The process involves checking out a `nixpkgs` commit on the selected
date, updating the [R]{.proglang} package set using Posit CRAN and Bioconductor
snapshots, backporting fixes, and ensuring popular packages work on both
x86-linux (including WSL2) and aarch64-darwin (Apple Silicon). These changes are
committed to a dated branch in `rstats-on-nix/nixpkgs`.
